``Interestingness'' in Mathematical Theory Exploration

### 2018-09-12 PhDThesis/litreview_staticanalysis.tex
Thesis changes (e.g. lit review)

Software for analysing and improving code can be roughly divided into two
categories. Those which contribute to the running of a program, such as
(optimising) compilers, linkers and interpreters must produce correct,
predictable, desirable results in a timely manner; for example, compiler
optimisations should not alter program semantics, should provide improvements in
the majority of cases (i.e. they should not be \emph{pessimisations}) and under
which they apply should be predictable enough to prevent spurious performance
regressions when making seemingly innocuous refactorings. Optimisations in
ahead-of-time (AOT) compilers must be fast enough to run during every build, and
should have a large enough impact on the generated code's performance that they
``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
ideals become hard requirements, since any runtime transformation which doesn't
pay for itself is, by definition, a pessimisation.

The other category is that of ``standalone'' tools, which are run separately to
building or executing the target code, and whose results do not contribute
directly to the code's output. These include tools for testing, profiling,
verification, standards compliance, etc. Our application of theory exploration
to software libraries fits into this latter category, whose constraints are not
as restricting as those applying to tools like compilers. The only real
restrictions on our tools are that their resource requirements are acceptable to
users, and that output is sufficiently high quality to be worth waiting for and
sifting through.

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used or
``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

A popular linter for Haskell code is \texttt{HLint}~\cite{mitchell2014hlint},
which focuses mostly on simplifying code; for example by spotting use-cases for
common library functions, e.g. replacing a pattern match \texttt{case A of \{
  Nothing -> B; Just C -> D C; \}} with the function call \texttt{maybe B D A}.
Linting can be seen as the inverse of theory exploration: rather than inferring
general relationships from repeated patterns in the code, linting begins with
known relationships and looks through the code for specific instances. Whilst
linting is a useful programming aid, and fast enough to apply regularly (e.g. as
a background process in a code editor), it is not an open-ended process: once
the predetermined patterns are exhausted, no more improvements will be found.

If fast but limited linters live at one end of a code improvement spectrum, at
the other end we find \emph{superoptimisers}~\cite{massalin1987superoptimizer}.
These treat the target program as a specification, and perform an open-ended
search through the space of all programs (in some specific language, usually a
subset of machine code which is free of branches and side-effects) to find all
possible implementations. Superoptimisers need some way to check program
equivalence, such as a test suite (to quickly eliminate incorrect programs) and
a theorem prover. Proving equivalence of machine code programs is helped by the
uniform, finite nature of their domain (e.g. a fixed set of 32bit registers).
Programs in higher level languages, like Haskell, may have unlimited input
domains (say, the natural numbers), which makes effective testing more difficult
and proofs potentially undecidable. A decidable method of comparison is also
required, to choose which of two equivalent programs is ``better''. This is
typically a cost model for the language, with a trivial example being program
length (i.e. assuming all instructions have the same cost).

>>>>

Software for analysing and improving code can be roughly divided into two
categories. Some contribute to the running of a program, such as (optimising)
compilers, linkers and interpreters. These must produce correct, predictable,
desirable results in a timely manner. For example, compiler optimisations should
not alter program semantics, should provide monotonic improvements in the
majority of cases (i.e. they should not be \emph{pessimisations}) and conditions
under which they apply should be predictable enough to prevent spurious
performance regressions when making seemingly innocuous
refactorings~\cite{robison2001impact}. Optimisations in ahead-of-time (AOT)
compilers must be fast enough to run during every build, and should have a large
enough impact on the generated code's performance that they
``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
ideals become hard requirements, since any runtime transformation which doesn't
pay for itself is, by definition, a pessimisation.

The other category is that of ``standalone'' tools, which are run separately to
building or executing the target code, and whose results do not contribute
directly to the code's output. These include tools for testing, profiling,
verification, standards compliance, etc. Our application of theory exploration
to software libraries fits into this latter category, whose constraints are not
as stringent as those applying to tools like compilers. The only real
restrictions on our tools are that their resource requirements are acceptable to
users, and that output is sufficiently high quality to be worth waiting for and
sifting through.

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used, or
``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

A popular linter for Haskell code is \textsc{HLint}~\cite{mitchell2014hlint},
which focuses mostly on simplifying code; for example by spotting use-cases for
common library functions like \texttt{null x} instead of \texttt{length x == 0}.
Linting can be seen as the inverse of theory exploration: rather than inferring
novel, general relationships from patterns found in the code, linting begins
with known relationships and looks through the code for specific instances.
Whilst linting is a useful programming aid, and fast enough to apply regularly
(e.g. as a background process in a code editor), it is not an open-ended
process: once the predetermined patterns are exhausted, no more improvements
will be found.

If fast but limited linters live at one end of a code improvement spectrum, at
the other end we find \emph{superoptimisers}~\cite{massalin1987superoptimizer}.
These treat the given code as a specification, and perform an open-ended search
through the space of all programs (in some specific language, usually a subset
of machine code which is free of branches and side-effects) to find the optimal
implementation according to some quality measure or cost model, such as
instruction count. Superoptimisers need some way to check program equivalence,
such as a test suite (to quickly eliminate incorrect programs) and/or a theorem
prover. Proving equivalence of machine code programs is helped by the uniform,
finite nature of their domain (e.g. a fixed set of 32bit registers). Programs in
higher level languages, like Haskell, may have unlimited input domains (say, the
natural numbers), which makes effective testing more difficult and proofs
potentially undecidable. Measuring and comparing efficiency of implementations
is also more complicated in such languages, potentially requiring some form of
cost semantics to derive an algebraic cost
formulation~\cite{danner2015denotational} and empirical testing to determine the
constants.
